---
title: "Simple Kriging"
author: "Yuting Fan"
date: "2022-10-03"
output: html_document
---

One way to build a kriging model in R is using 'krige' function in 'gstat' package. Here is an example to show the code of building simple kriging model corresponding to the theoretical formula. 

Consider a spatial process $Y(s)$ with known $C_Y(s,u)=cov(Y(s),Y(u))$ and known $E(Y(s))$. The goal is to predict $Y(s)$ based on $\boldsymbol{Z}=(Z(s_1),\dots,Z(s_n))^{'}$ observed according to $Z(s_i)=Y(s_i)+\epsilon(s_i)$. 

The predictors of $Y(s_0)$ is $\hat{Y}_{sk}(s_0)=\mu(s_0)+\boldsymbol{c}^{'}\Sigma_{Z}^{-1}(\boldsymbol{Z}-\mu)$, where $\boldsymbol{c}=(C_Y(s_0,s_1),\dots,C_Y(s_0,s_n))^{'}$ and $\Sigma_Z$ is the $n \times n$ matrix whose $(i,j)-th$ element is $C_Y(s_i,s_j)$, and the mean-squared prediction error is $\sigma^{2}_{sk}(s_0)=C_Y(s_0,s_0)-\boldsymbol{c}^{'}\Sigma_{Z}^{-1}\boldsymbol{c}$.

Here, I assume $C_Y(s,u)$ is a Matern covariance function with parameters specified in the following chunk. For simplicity, $E(Y(s))$ is assumed a constant 0. 

```{r, message=FALSE, warning=FALSE}
library(knitr)
library(geoR)
library(gstat)
library(sp)
```

# Create data

Set the mean function and the parameters of covariance function.

```{r}
# parameters of spatial process
sill = 5 # sigma^2
range = .25
smoothness = 0.5
nugget = 0
mean = 0
```

Generate a realization of Gaussian random fields with Matern covariance function. The location is $s \in [0,5]\times[0,5]$.

```{r, includes=FALSE}
# generate observed data
n = 5
r = grf(n, 
        grid = "irreg", 
        xlims = c(0, 5), ylims = c(0, 5),
        nsim = 1, 
        cov.model = "matern",
        cov.pars = c(sill, range),
        kappa = smoothness, nugget = nugget,
        mean = mean)

data = data.frame(response = r$data,
                  x = r$coords[,1],
                  y = r$coords[,2])
coordinates(data) = ~x+y

# location for prediction
pred_loc = data.frame(x = c(3),
                      y = c(3))
newdata = data.frame(x = pred_loc$x,
                     y = pred_loc$y)
coordinates(newdata) = ~x+y
```

The number of observed locations is 5. The location to predict is (3,3).

```{r}
df = data.frame(x = r$coords[,1],
                y = r$coords[,2],
                value = r$data)
kable(df, caption = "Observation")
kable(pred_loc, caption = "To predict")
```

# Build the model using krige

```{r}
# build kriging model using gstat package
m <- vgm(psill = sill, "Mat", range = range, nugget = nugget, kappa = smoothness)
model = krige(response ~ 1, data, newdata = newdata, model = m, beta = 0)
```

# Compute from scratch

Compute the predictor and MSPE of simple kriging acoording to the observation $Z$.

```{r}
Z = as.matrix(data@data)

# compute the distance
distmat = as.matrix(dist(x = rbind(data@coords, newdata@coords), method = "euclidean", diag = TRUE, upper = TRUE))
distmat_obs = distmat[1:n, 1:n]
distmat_pred = distmat[1:n,n+1, drop = FALSE]

# \Sigma_Z, c, and w
cov_funct = GPvecchia::MaternFun(distmat_obs, covparms = c(sill, range, smoothness)) # \Sigma_Z
cov_vect = GPvecchia::MaternFun(distmat_pred, covparms = c(sill, range, smoothness)) # c
w = solve(cov_funct)%*%cov_vect

# predictor and mspe
pred = t(w)%*%(Z - mean)
var = sill - t(cov_vect)%*%solve(cov_funct)%*%cov_vect
```

# Compare the results

```{r}

result = data.frame(prediction = c(model$var1.pred, pred),
                    mspe = c(model$var1.var, var))
rownames(result) = c("gstat::krige", "scratch")
kable(result)
```

